State of the Art in Scientific Workflow Systems

### 2014-05-23 paper.tex
calibrate

\section{Introduction}
\label{sec:intro}
\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE from here on)
are, in principle similar to the traditional programming environments
(programming language, interpreter or compiler, and runtime) albeit with a
larger scale of data and a higher abstraction of computation. The similarity is
such that scripting languages are sometimes advocated for programming at this
level. However, traditional scripts do not deal with parallelism and dataflow
aspects of workflow paradigm which makes them a strong contenders for the
purpose of high-level application programmers.
%
%Add more content here to beef up the context.
%
% Goals and benefits of a workflow management system
The goals of a workflow management system for scientific computing are as
follows:

\begin{enumerate}
\item \textbf{Bulk Work.}
Workflows are important because they enable to perform computational work in
bulk. This could be bulk data processing or running multiple distinct
processing tasks on a single set of data. In addition to carrying out such
work, workflow environments provide oppurtunities for giving structure and
order to a set of tasks and often with sufficient expressivity, choices and
selections to carry out a bulk of work selectively. Some of the examples of
such work are \emph{parameter sweeps} where tasks are run over a series of
parameter values, \emph{ensemble studies} where tasks are carried out over
discrete groups of parameters over a space.

\item \textbf{User Interaction.} Adaptation to new methods of computing is not
easy for domain science users. Such users often need special assistance in
order to translate their computation requirements into concrete computing
entities.  SWEs are an ideal platform enabling users to perform many tasks
conveniently.  These tasks include visually formulating a task, composing
and marshalling of workflow, browsing and selecting data, visualizing and
sharing results.

\item \textbf{Resource Access and Utilization.} Computational platforms and
access modalities change rapidly with upgradation of existing technologies and
adaptation of new techniques. Clouds are the latest examples of new techniques
of computational platform access while GPGPUs are examples of upgradation of
existing processing environments. Additionally, with the advent of fast and
dedicated networks such as ESNet and federated visualization resources, remote
visualization is gaining prominence among scientific users. SWEs play
important role as platforms to give transparent and hassle free access to such
novel and multiple environments to users by hiding the technicalities of
access methods and usage and configuration related nuances.

@@

\item \textbf{Provenance and MetaData.} Historical data and metadata has been
crucial entities in many research fields. They help connecting results from
a multitude of research aspects. A provenance trace of computation can help
checkpoint methods and reproduce and modify partial computation. Automated
observations of deviations in scientific phenomena from provenance data is one
of the key triggers for new scientific discoveries. Fields of studies involving
genomic and protiens are examples of domains where provenance and metadata
plays key roles. SWEs can be integrated or extended with metadata and
provenance techniques rendering powerful computing and traceability features.

\item \textbf{Sharing and Collaboration.} Workflow systems are often targetted
towards a particular community of users. Furthermore, dissemination of methods and results
to communities and collaborators are some of the activities that SWEs make
convenient for end-users. For instance, the Galaxy workflow system is targetted
specifically for the bioinformatics users encouraging them to collaborate over
problems and share results. Another such example is Taverna which is geared
towards sharing and collaborative workflow oriented scientific computation
enabling users to discover new services introduced to a community. Sharing and
collaboration of applications and related data have become ubiquitious with the
rise of social web. Initiatives such as
\emph{myExperiment}~\cite{deroure-goble-etal:2008} and recently
\emph{gitHub}~\cite{github-web} have enabled easy sharing and collaboration of
reproducible workflows on the social web.

@@

The above described goals also form a basis for the need for a SWE. The need
for the use of SWE can arise as a result of one or more of such requirements.
Traditional programming platforms are not suitable to address these
requirements as they operate at a different level of abstraction. SWEs differ
in many respects among themselves as well as with respect to traditional
programming environments. The following list identifies the major differences
between traditional and the SWE programming paradigm:

\begin{enumerate}
%

\item \textbf{Scalability and Optimization.} SWEs are more likely to involve
extensive usage of distributed resources than conventional programs. Owing
to the large-scale experiments, for which scientific users turn to automation,
their demands for computational power often exceeds that of a single personal
computer. These properties make them ideal candidates for a large-scale
distributed computing environment. The case for suitability of SWEs on
large-scale distributed infrastructures has been well-argued upon
in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to
ensure scalable and optimized performance is vital. Additionally, most SWEs are
capable of executing workflows at small scale with minimal tuning. This
multi-scalability enables quick and easy conceptualization and prototyping of
new workflows.
%

\item \textbf{Data Management.} The scale of data transfer is much higher in
SWEs from one execution site to others as well as among storage resources while
it is lower (ranging from variables to small data structures) in programs.
Owing to such scales, applications often need access to special data movement
and access protocols such as GridFTP, Amazon S3, NAS, etc. SWEs are often
shipped with capabilities to handle such protocols.
%

\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
%TODO: Add content about new and emerging Distributed Computing
    %Infrastructures: Clouds, GPUs and exascale supercomputers
Building SWE's interface to distributed systems is a complex task. The
distributed nature of resource utilization and asynchronous execution of
workflow requires a set of interdependent parameters to be configured
correctly. Modern computing infrastructures such as Clouds, GPUs and
supercomputers with hybrid nodes pose new challenges of interface development.
As SWEs involve running many tasks over an infrastructure it becomes crucial to
interface with these systems such that the tasks are scheduled in an optimal
manner. Many SWEs implement
\emph{pilot-jobs}~\cite{montagnat-glatard-etal:2010,ucc2011} in order to
reserve resources via pseudo-tasks that act as placeholders for real tasks.
Additionally, decision to manage working data storage on disc, memory and cache
are some of the distinctions between a traditional programming and workflow
platforms.

\item \textbf{Expressiveness.} Programming languages are inherently control
flow oriented where data flow is a secondary aspect. In fact, the
programmer must explicitly make sure that correct the inputs are available
to instructions and are in the right format. A program proceeds as the
execution of its statements from beginning to end and completes its
execution in a predefined sequence. However, in SWEs, the flow is often
determined by the data availability for execution. This model of computation is
known as dataflow computing. A workflow thus does not follow a fixed order of
execution. Different execution sequences are possible which are dominated by
dataflow. The unit of work in SWEs is at a higher level of granularity. A unit
of execution is coarse-grained, and often termed as ``processor'', ``task'',
``service'' or ``actor'' in the SWE community. This notion of high-level
granularity drives the depth to which a workflow can be expressed in a given
language form. In the rest of the paper, we refer to a unit of execution of a
workflow system as \emph{processor}. On the other hand, a unit of execution in
most programming languages is a much simpler and finer executable statement
that accomplishes a small computational task. This presents a whole new set of
requirements for expressing the processors and how they should interact with
computational resources and operate upon data.

%
\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain
experts to utilize the computational power. This makes it essential for
them to have a friendly, and intuitive user-interface for novice and
non-expert users. For this reason, SWE user interfaces are dominated by the
visual paradigm of programming. On the other hand, traditional programs are
most often written by expert programmers. Thus, they have a text-based
composition environment albeit with visually assisting programming environments
such as IDEs. However, they are seldom to the extent that a non-expert user can
graphically compose programs. While a graphical interface enhances usability, a
text-based workflow environment can arguably give better control to an
experienced user. Consequently, many workflow systems offer a dual interface of
graphical and text medium to compose workflows.

@@

We concede that the above is not an exhaustive list of differences in features.
However, it is reasonably inclusive and we believe it covers the breadth of
aspects we cover in the current work. We do not include other interesting SWE
features such as application predictability and dynamic restructure of DAGs in
this analysis as they are out of scope of the objectives of the current paper.

@@

\subsection{Workflow}
Figure~\ref{fig:wf} shows a typical workflow in the form of a Directed Acyclic
Graph (DAG). For the purpose of this text, we define a workflow as a
computation with at least two discrete tasks either forming distinct
computational stages or running for two or more distinct datasets as
parameters. Consequently, a one-off task or running a task for a single copy of
data is \emph{not} a workflow. For the rest of this paper, we will use the above
definition of a workflow.
%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=4cm]{figures/workflow}
\caption{A typical workflow depicted as a Directed Acyclic Graph (DAG) of tasks coordinated by data or control flow and illustrating data and task parallelism}
\label{fig:wf}
\end{center}
\end{figure}
%
\subsection{Architecture of a Generic SWE}
Figure~\ref{fig:wfms} shows a high-level architecture of a generic SWE. It shows key
components and their interaction in a SWE. These components are the focus of
interest for the present work. A \textit{user interface} interacts with the
\textit{workflow system} using commands and messages on the inside while with
user on the outside. The \textit{workflow language} becomes an intermediate
representation for communication between user interface and the
\textit{workflow system}. The workflow system acts as a compiler/interpreter
and a high-level scheduler for the processors and interfaces with the underlying
\textit{distributed resources}. It also monitors the workflow progress and
collects the intermediate and final results.

%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{figures/workflow_interfaces_and_language}
\caption{A generic SWE scheme showing the relationship and interaction among important components}
\label{fig:wfms}
\end{center}
\end{figure}
%
SWEs are used as tools to benefit from the underlying large-scale computing
infrastructure such as clusters, supercomputers, clouds, grids. Such
environments are in use today for production level work in different scientific
domains worldwide. They do the basic task of linking inter-related processors
of a multi-staged scientific computation. However, they differ in the features
they provide based upon the scientific domain they offer service for, and the
computational model they follow. Often, one of the main reasons for such
difference is the specialized components and requirements for a particular
domain. Owing to this fact, a given workflow system dominates a particular area
of science (eg. astronomy, biology, etc.). Nonetheless, there exist a plethora
of SWEs suitable for general purpose workflow style computations.

\section{Related Work}
Surveys and studies on workflows similar to the present one have been done in
the past. Yu and Buyya \cite{yu-buyya:2005} proposed a detailed taxonomy of
SWEs, studying many existing systems for distributed computing. This taxonomy
is based, to a large extent on the capabilities of the workflow enactors. It
provides only a limited insight to the SWE aspects considered in the present
paper, especially the workflow languages expressiveness and SWEs interactions
with the distributed computing infrastructures. A more recent work by Goderis \cite{goderis-brooks-etal:2009} studies a topic more closely
related to the workflow language expressiveness: it compares the models of
computations of various workflow directors implemented in Kepler
\cite{ludascher-altintas-etal:2005a} with the perspective of enabling
composition of different models in sub-workflows of a global orchestration.
This work illustrates how different enactors designed for different application
needs have different expressiveness power. Deelman et. al.
\cite{deelman-overview} present a high-level overview of the scientific
workflow features and capabilities including SWE composition, representation
and resource mapping capabilities. The recent survey from Stratan~\cite{stratan-iosup-etal:2008} about the workflow engine performance takes into
account the latencies involved during the enactment mechanisms. A small survey
and analysis of file-access characteristics for data-intensive workflows has
been done by Shibata~\cite{Shibata}. Early surveys on workflow
languages and suitability of general purpose languages to workflows is
conducted in \cite{generalpurpose}. These surveys were conducted before a few
years and in the meantime SWEs have come a long way in terms of capabilities
and enactment approaches. New techniques such as the map/reduce framework
\cite{mapreduce} have been introduced as a promising new approach to workflow
enactment over distributed infrastructures. Several issues put forth by such
surveys such as customized service wrapping and enactment have been adequately
addressed over the
years~\cite{glatard-montagnat-etal:2008a,rojasbalderrama-montagnat-etal:2010}.

Another well-studied feature is the workflow languages. Many different
languages have been considered within the SWE community, from raw Directed
Acyclic Graphs (DAGs) of computational processes such as CONDOR DAGMan
\cite{dagman} to abstractions for parallel computations such as Petri nets
\cite{alt-hoheisel:2005},
data-driven languages such as Scufl \cite{turi-missier-etal:2007} and scripting
languages such as Swift script \cite{swift}. Glatard's PhD thesis
\cite{glatard:2007} presents a taxonomy of SWE languages classifying them into
functional, graph, service-oriented and executable representations. A detailed
classification and taxonomy of parallelism patterns in Grid Workflows has been
presented by Pautasso and Alonso in \cite{pautasso1}. This classification is
more generic while the one presented in Glatard's thesis has emphasis on
service based workflows.  Each of these approaches can be defended through some
aspects well covered in their design: DAGs are convenient for scheduling
\cite{Malewicz:2007,hall-rosenberg-etal:2007}, Petri nets can be used to detect
properties such as potential deadlocks, data-driven languages ease the
description of application logic for non-expert users and scripting is
extensively used by programmers for prototyping, etc.

DAGs are by far, the most popular abstraction of the way workflows are
perceived and visualized. The MA
DAG~\cite{caron-desprez:2005},\km{could not find any recent paper on MA DAG} engine is one example of a DAG generator engine. MA
DAG has evolved from a regular DAG generator to a more abstract DAG generator
laced with control structures. MA DAG benefits from the workflow scheduling
mechanisms included in the middleware. Due to the impossibility to produce
complete DAGs owing to the presence of control structures such as conditionals
and loops, MA DAG generate partial DAGs greedily: a sub-DAG, as complete as
possible, is produced as soon as possible. Despite the presence of many
different approaches to workflow systems, new requirements
continue to pose workflow challenges. Some of the current challenges facing SWEs are:
\begin{enumerate}
    \item \textbf{Multisite execution.} Interfacing to existing computation and
        storate infrastructure is not trivial, not to mention the new and
        unforeseen access and compute modes. SWEs are yet to formulate a
        generalized interface towards the major platforms such that they will
        run on them out of the box.

@@

    \item \textbf{Interoperability.} Interoperable workflows will benefit to a
        broad audience of users. Many efforts at this are made in the recent
        past, such as by the consurtium projects and by individual players.
        However, a truly interoperable set of standards for workflows are yet
        to emerge.
\end{enumerate}
\section{Objectives and Organization}
% What are the goals of this work?
% How is this paper supposed to help:
% -- Existing scientific users who have not adopted workflows
% -- Existing users who have adopted workflows
% -- New users
% -- Experts in the field

@@

\begin{enumerate}
  \item \textbf{Educate.} With more and more science being done with the help of
      computational methods, SWEs can add significant value to the process.
      They can accelerate the process, enable community driven science, drive
      the development at a manageable level of abstraction for users.
      However, scientific users are often not experienced programmers or
      compute savvy. This work will serve as a useful starting guide to
      educate science users of the available tools and techniques with
      features that might be suitable for their methods.

 \item \textbf{Catalog.} We study a number of important features as offered by
     contemporary SWEs. This will help catalog and collect the SWEs by their
     offerings and identify mechanisms and needs for new and existing features
     in one place.  This will also help identify current trends and approaches
     to computation in scientific communities.

 \item \textbf{Application.} We identify the applications and application
     classes well adapted to the workflow systems and are benefitting by them.
     This will help readers relate to the application they are associated to.
     We hope that readers can make a choice of which SWE to use if there is an
     adaptation scope.

 \item \textbf{Evolution.} We comment on the evolution of the existing workflow
     systems and the possible current and future development directions which
     will help readers be aware of the general development direction of the
     SWEs.

The rest of this paper is organized as follows: Section
\ref{sec:overview} presents an overview of the key SWEs studied in this work.
Section \ref{sec:scal} discusses the optimization features and resulting
scalable performance as observed in the SWEs studied. Section \ref{sec:data}
studies the data management and description aspects of the SWEs. Section
\ref{sec:inter} presents a study of SWEs approaches to interface the
large-scale computing infrastructures. Section \ref{sec:lang} details the
language features and expressiveness of the SWEs studied. Section
\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
\ref{sec:sum} presents a summary and a classification of all the aspects of
SWEs studied in this work. Throughout the paper, we also present tables
summarizing the aspects of SWEs considered in respective sections. It was not
possible to obtain an exhaustive information about all the SWEs considered in
this work. We rely on the following sources for our information: 
\begin{enumerate}
    \item \textbf{Publications.} Peer reviewed publications in journals, conferences and workshops.
    \item \textbf{Technical Reports.} Technical reports and theses.
    \item \textbf{Web Presence.} Documentation and other information from the web pages hosting the SWE.
    \item \textbf{Experience.} Where possible, we tried features of the SWEs by installing them on the available systems. 
\end{enumerate}

Some features, such as extendability of SWE strongly depend on the extension
points designed in the code; consequently, we consider it present in only those
SWEs that have an explicit documented description or demonstration of the given
feature.

\section{Overview of the Workflow Environments} \label{sec:overview}
In this section we present an overview of the SWEs studied.  The paragraph
headers are organized by the name of the SWE followed by the name of internal
representation language use and support separated by a forward slash (/).  The
cases where an intermediate language is not specifically defined or conversely
where there are multiple systems supporting a generic representation are
denoted by an asterisk (*) or a hyphen (-) at the respective places.
\paragraph{Taverna/scufl} The Taverna Workflow Management System
\cite{taverna2013} is an open source scientific dataflow manager
developed by the mygrid~consortium in the UK.
%
Taverna includes a GUI-based rich-client workbench for workflow design and a
workflow enactment runtime. Taverna can enact workflows expressed in the scufl
language. Scufl is a near Turing complete dataflow language with well-defined
semantics \cite{turi-missier-etal:2007,glatard-montagnat:2008}.

>>>>

\section{Introduction}
\label{sec:intro}
\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE)
are, in principle similar to the traditional programming environments
(programming language, interpreter or compiler, and runtime) albeit with a
larger scale of data and a higher abstraction of computation. The similarity is
such that scripting languages are sometimes advocated for programming at this
level. However, traditional scripts do not deal with parallelism and dataflow
aspects of workflow paradigm which makes them a strong contenders for the
purpose of high-level application programming. Computation in science is
dominated by computation for data analysis via one or more mathematical or
scientific model execution. These models of computation are composed of
multiple stages where each stage executes a program requiring input parameters
and data while producing results in the form of data which in turn acts as
input to the next stage. SWEs can ease these activities by supporting the
expression, execution and automation.
% Goals and benefits of a workflow management system
The goals and benefits of an SWE are as follows:

\begin{enumerate}
\item \textbf{Bulk Work.}
Workflows are important because they enable to perform computational work in
bulk. This could be bulk data processing or running multiple distinct
processing tasks on a single set of data. In addition to support such
work, workflow environments provide oppurtunities for giving structure and
order to a set of tasks and often with sufficient expressivity, choices and
selections to carry out a bulk of work selectively. Some of the examples of
this approach are \emph{parameter sweeps} where tasks are run over a series of
parameter values, \emph{ensemble studies} where tasks are carried out over
discrete groups of parameters over a space.

\item \textbf{User Interaction.} Adaptation to new methods of computing is not
easy for domain science users. Such users often need special assistance in
order to translate their computation requirements into concrete computing
entities. SWEs are an ideal platform enabling users to perform many tasks
conveniently. These tasks include visually formulating an analysis, composing
and marshalling of processes, browsing and selecting resulting data,
visualizing and sharing final results.

\item \textbf{Resource Access and Utilization.} Computational platforms and
access modalities change rapidly with upgradation of existing technologies and
adaptation of new techniques. Clouds are an example of new techniques
of computational platform access while GPGPUs are example of upgradation of
existing processing environments. Additionally, with the advent of fast and
dedicated networks such as ESNet~\cite{esnet} and federated remote
visualization resources are gaining prominence among scientific users. SWEs
play important role as platforms to give transparent and hassle free access to
such novel and multiple environments to users by hiding the technicalities of
access methods and usage and configuration related nuances.

@@

\item \textbf{Provenance and MetaData.} Historical data and metadata has been
crucial entities in many research fields. They help connecting results from
a multitude of research points of view. A provenance trace of a computation can
help checkpoint methods and reproduce and modify partial computation. Automated
observations of deviations in scientific phenomena from provenance data is one
of the key triggers for new scientific discoveries. Fields of studies involving
genomic and protiens are examples of domains where provenance and metadata
plays key roles. SWEs can be integrated or extended with metadata and
provenance techniques rendering powerful computing and traceability features.

\item \textbf{Sharing and Collaboration.} SWEs are often targetted
towards a particular community of users. Dissemination of methods and results
to communities and collaborators are some of the activities that SWEs make
convenient for end-users. For instance, the
Galaxy~\cite{blankenberg-vonkuster-etal:2010} workflow system is targetted
specifically for the bioinformatics users encouraging them to collaborate over
problems and share results. Another such example is Taverna~\cite{taverna2013}
which is geared towards sharing and collaborative workflow oriented scientific
computation enabling users to discover new services introduced to a community.
Sharing and collaboration of applications and related data have become
ubiquitious with the rise of social web. Initiatives such as
\emph{myExperiment}~\cite{deroure-goble-etal:2008} and recently
\emph{gitHub}~\cite{github-web} have enabled easy sharing and collaboration of
reproducible workflows on the social web.

@@

The above described goals and benefits also form a basis for the need for a
SWE. The need for the use of SWE can arise as a result of one or more of such
requirements. Traditional programming platforms are not suitable to address
these requirements as they operate at a different level of abstraction. SWEs
differ in many respects among themselves as well as with respect to traditional
programming environments. Consequently, a choice of the suitable SWE must be
made by a researcher. The following list identifies the major differences
between traditional and the SWE programming paradigm:

\begin{enumerate}
%
\item \textbf{Scalability and Optimization.} SWEs are more likely to involve
extensive usage of distributed resources than conventional programs. Owing
to the large-scale experiments, for which scientific users turn to automation,
their demands for computational power often exceeds that of a single personal
computer. These properties make them ideal candidates for a large-scale
distributed computing environment. The case for suitability of SWEs on
large-scale distributed infrastructures has been well-argued upon
in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to
ensure scalable and optimized performance is vital. Additionally, most SWEs are
capable of executing workflows at small scale with minimal tuning. This
multi-scalability enables quick and easy conceptualization and prototyping of
new workflows.
%
\item \textbf{Data Management.} The scale of data transfer is much higher in
SWEs from one execution site to others as well as among storage resources while
it is lower (ranging from variables to small data structures) in programs.
Owing to such scales, applications often need access to special data movement
and access and movement protocols such as GridFTP, Amazon S3, NAS, etc. SWEs are often
shipped with capabilities to handle such protocols.
%

\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
Supporting an interface to a distributed system is a complex task from the
point of view of an SWE. The distributed nature of resource utilization and
asynchronous execution of workflow requires a set of interdependent parameters
to be configured correctly. Modern computing infrastructures such as Clouds,
GPUs and supercomputers with hybrid nodes (CPU + GPGPU) pose new challenges of
interface development.  As SWEs involve running many tasks over an
infrastructure it becomes crucial to interface with these systems such that the
tasks are scheduled in an optimal manner. Many SWEs implement
\emph{pilot-jobs}~\cite{montagnat-glatard-etal:2010,ucc2011} in order to
reserve resources via pseudo-tasks that act as placeholders for real tasks.
Additionally, decision to manage working data storage on disc, memory and cache
are some of the distinctions between a traditional programming and workflow
platforms in the context of distributed and scalable computation.

\item \textbf{Expressiveness.} Programming languages are inherently control
flow oriented where data flow is a secondary aspect. In fact, the
programmer must explicitly make sure that correct inputs are available
to instructions and are in the right format. A program proceeds as the
execution of its statements from beginning to end and completes its
execution in a predefined sequence. However, in SWEs, the flow is often
determined by the data availability for execution. This model of computation is
known as dataflow computing. A workflow thus does not follow a fixed order of
execution. Different execution sequences are possible which are dominated by
dataflow. The unit of work in SWEs is at a higher level of granularity. A unit
of execution is coarse-grained, and often termed as ``processor'', ``task'',
``service'' or ``actor'' in the SWE community. This notion of high-level
granularity drives the depth to which a workflow can be expressed in a given
language form. In the rest of the paper, we refer to this unit of execution of a
workflow system as \emph{processor}. On the other hand, a unit of execution in
most programming languages is a much simpler and finer executable statement
that accomplishes a small task. This presents a whole new set of
requirements for expressing the processors and how they should interact with
computational resources and operate upon data.

%
\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain
experts to utilize the computational power. This makes it essential for
them to have a friendly, and intuitive user-interface for novice and
non-expert users. For this reason, SWE user interfaces are dominated by the
visual paradigm of programming. On the other hand, as the sophistication of
expression and user expertise increases, the visual environment temds to get
limited in capabilities it can offer. Thus, most SWEs have a text-based
composition environment albeit with visually assisting programming environments
such as IDEs. However, these textual environments are seldom to the extent that
a non-expert user can graphically compose programs. While a graphical interface
enhances usability, a text-based workflow environment can arguably give better
control to an experienced user. Consequently, many workflow systems offer a
dual interface of graphical and text medium to compose workflows.

@@

We concede that the above is not an exhaustive list of differences in features.
However, it is reasonably inclusive and we believe it covers the breadth of the
aspects we discuss in the current work. We do not include other interesting SWE
features such as application predictability in this analysis as they are out of
scope of the objectives of the current paper.

@@

\subsection{Workflow}
Figure~\ref{fig:wf} shows a typical workflow in the form of a Directed Acyclic
Graph (DAG). Execution typically begins with the input data being provided to
the first task and ends at the results being delivered at the end of the last
task. Tasks are represented by boxes and data/control flow between the task is
represented as arrows. Note that in cases where tasks are connected with pure
control flow, it can be simulated by passing token data between tasks.

For the purpose of this text, we define a workflow as a specification of
computation with at least two discrete tasks either forming distinct
computational stages or running for two or more distinct datasets as
parameters. Consequently, a one-off task or running a task for a single copy of
data is \emph{not} a workflow.

For the rest of this paper, we will use the above definition of a workflow.
%

\begin{figure}[htb]
\begin{center}
\includegraphics[width=4cm]{figures/workflow}
\caption{A typical workflow depicted as a Directed Acyclic Graph (DAG) of tasks coordinated by data or control flow and illustrating data and task parallelism}
\label{fig:wf}
\end{center}
\end{figure}
%
\subsection{Architecture of a Generic SWE}
Figure~\ref{fig:wfms} shows a high-level architecture of a generic SWE. It shows key
components and their interaction in an SWE. These components are the focus of
interest for the present work. A \textit{user interface} interacts with the
\textit{workflow engine} using commands and messages on the inside while with
user on the outside. The \textit{workflow language} becomes an intermediate
representation for communication between user interface and the
\textit{workflow engine}. The workflow engine acts as a compiler/interpreter
and a high-level scheduler for the processors and interfaces with the
underlying \textit{distributed resources}. It resolves data dependencies and
optimizes the flow at runtime and monitors the workflow progress and collects
the intermediate and final results.

%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{figures/workflow_interfaces_and_language}
\caption{A generic SWE scheme showing the relationship and interaction among important components}
\label{fig:wfms}
\end{center}
\end{figure}
%
SWEs are used as tools to benefit from the underlying large-scale computing
infrastructure such as clusters, supercomputers, clouds, and grids. Such
environments are in use today for production level work in different scientific
domains worldwide. They do the basic task of linking inter-related processors
of a multi-staged scientific computation. However, they differ in the features
they provide based upon the scientific domain or user community they offer
service for, and the computational model they follow. Often, one of the main
reasons for such difference is the specialized components and requirements for
a particular domain. Owing to this fact, a given workflow system dominates a
particular area of science (eg. astronomy, biology, etc.). Nonetheless, there
exist a plethora of SWEs suitable for general purpose workflow style
computations.

DAGs are by far, the most popular abstraction of the way workflows are
perceived and visualized. The MA DAG~\cite{caron-desprez:2005}, engine is one
example of a DAG generator engine. MA DAG has evolved from a regular DAG
generator to a more abstract DAG generator laced with control structures. MA
DAG benefits from the workflow scheduling mechanisms included in the
middleware. Due to the impossibility to produce complete DAGs owing to the
presence of control structures such as conditionals and loops, MA DAG generate
partial DAGs greedily: a sub-DAG, as complete as possible, is produced as soon
as possible. Despite the presence of many different approaches to workflow
systems, new requirements continue to pose workflow challenges. Some of the
current challenges facing SWEs are:

\begin{enumerate}
\item \textbf{Multisite execution.} Interfacing to existing computation and
storate infrastructure is not trivial, not to mention the new and
unforeseen access and compute modes. SWEs are yet to formulate a
generalized interface towards the major platforms such that they will
run on them out of the box.

@@

\item \textbf{Interoperability.} Interoperable workflows will benefit to a
broad audience of users. Many efforts at this are made in the recent
past, such as by the consurtium projects and by individual players.
However, a truly interoperable set of standards for workflows are yet
to emerge.
\end{enumerate}

\section{Related Work}
Surveys and studies on workflows similar to the present one have been done in
the past. Yu and Buyya \cite{yu-buyya:2005} proposed a detailed taxonomy of
SWEs in the previous decade, studying many existing systems for distributed
computing. This taxonomy presented is based, to a large extent on the
capabilities of the workflow engines. It does provide insights to many of the
SWE aspects considered in the present paper, especially the workflow languages
expressiveness and SWEs interactions with the distributed computing
infrastructures in the early days of the compute Grids. A more recent work by
Goderis \cite{goderis-brooks-etal:2009} studies a topic more closely related to
the workflow language expressiveness: it compares the models of computations of
various workflow directors implemented in Kepler
\cite{ludascher-altintas-etal:2005a} with the perspective of enabling
composition of different models in sub-workflows of a global orchestration.
This work illustrates how different enactors designed for different application
needs have different expressiveness power. Deelman et. al.
\cite{deelman-overview} present a high-level overview of the scientific
workflow features and capabilities including SWE composition, representation
and resource mapping capabilities. A recent survey from
Stratan~\cite{stratan-iosup-etal:2008} about the workflow engine performance
takes into account the latencies involved during the enactment mechanisms. A
small survey and analysis of file-access characteristics for data-intensive
workflows has been done by Shibata~\cite{Shibata}. Early surveys on workflow
languages and suitability of general purpose languages to workflows is
conducted in~\cite{generalpurpose}. These surveys were conducted before a few
years and in the meantime SWEs have come a long way in terms of capabilities
and enactment approaches. New techniques such as the map/reduce framework
\cite{mapreduce} have been introduced as a promising new approach to workflow
enactment over distributed infrastructures. Several issues put forth by such
surveys such as customized service wrapping and enactment have been adequately
addressed over the
years~\cite{glatard-montagnat-etal:2008a,rojasbalderrama-montagnat-etal:2010}.

Another well-studied feature is the workflow languages. Many different
languages have been considered within the SWE community, from raw Directed
Acyclic Graphs (DAGs) of computational processes such as CONDOR
DAGMan~\cite{dagman} to abstractions for parallel computations such as Petri
nets~\cite{alt-hoheisel:2005}, data-driven languages such as
Scufl~\cite{turi-missier-etal:2007} and scripting languages such as Swift
script~ \cite{swift}. Glatard's PhD thesis~\cite{glatard:2007} presents a
taxonomy of SWE languages classifying them into functional, graph,
service-oriented and executable representations. A detailed classification and
taxonomy of parallelism patterns in Grid Workflows has been presented by
Pautasso and Alonso in~\cite{pautasso1}. This classification is more generic
while the one presented in Glatard's thesis has emphasis on service based
workflows.  Each of these approaches can be defended through some aspects well
covered in their design: DAGs are convenient for scheduling
\cite{Malewicz:2007,hall-rosenberg-etal:2007}, Petri nets can be used to detect
properties such as potential deadlocks, data-driven languages ease the
description of application logic for non-expert users and scripting is
extensively used by programmers for prototyping, etc.

\section{Objectives and Organization}
% What are the goals of this work?
% How is this paper supposed to help:
% -- Existing scientific users who have not adopted workflows
% -- Existing users who have adopted workflows
% -- New users
% -- Experts in the field

@@

\begin{enumerate}
\item \textbf{Educate.} With more science being done with the help of
computational methods, SWEs can add significant value to the process.  They can
accelerate the process, enable community driven science, drive the development
at a manageable level of abstraction for users.  However, scientific users are
often not experienced programmers or computation savvy. This work will serve as
a useful starting guide to educate science users and students of the available
tools and techniques with features suitable for their methods.

\item \textbf{Catalog.} We study a number of important features as offered by
the contemporary SWEs. This will help catalog and collect the SWEs by their
offerings and identify mechanisms and needs for new and existing features
in one place. This will also help identify current trends and approaches
to computation in scientific communities.

\item \textbf{Application.} We identify the applications and application
classes well adapted to the workflow systems and are benefitting by them.
This will help readers relate to the application they are associated to.
We hope that readers can make a choice of which SWE to use if there is an
adaptation similar in scope to the ones describe in this paper.

\item \textbf{Evolution.} We comment on the evolution of the existing workflow
systems and the possible current and future development directions which
will help readers be aware of the general development direction of the
SWEs.
\end{enumerate}

The rest of this paper is organized as follows: Section
\ref{sec:overview} presents an overview of the key SWEs studied in this work.
Section \ref{sec:scal} discusses the optimization features and resulting
scalable performance as observed in the SWEs studied. Section \ref{sec:data}
studies the data management and description aspects of the SWEs. Section
\ref{sec:inter} presents a study of SWEs approaches to interface the
large-scale computing infrastructures. Section \ref{sec:lang} details the
language features and expressiveness of the SWEs studied. Section
\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
\ref{sec:sum} presents a summary and a classification of all the aspects of
SWEs studied in this work.

Throughout the paper, we also present tables
summarizing the aspects of SWEs considered in respective sections. It was not
possible to obtain an exhaustive information about all the SWEs considered in
this work. We rely on the following sources for our information: 
\begin{enumerate}
    \item \textbf{Publications.} Peer reviewed publications in journals, conferences and workshops.
    \item \textbf{Technical Reports.} Technical reports and theses.
    \item \textbf{Web Presence.} Documentation and other information from the web pages hosting the SWE.
    \item \textbf{Experience.} Where possible, we tried features of the SWEs by installing them on the available systems. 
\end{enumerate}

Some features, such as extendability of an SWE strongly depend on the extension
points designed in the code; consequently, we consider it present in only those
SWEs that have an explicit documented description or demonstration of the given
feature.

\section{Overview of the Workflow Environments} \label{sec:overview}
In this section we present an overview of the SWEs studied. The paragraph
headers are organized by the name of the SWE followed by the name of internal
representation language use and support separated by a forward slash (/).  The
cases where an intermediate language is not specifically defined or conversely
where there are multiple systems supporting a generic representation are
denoted by an asterisk (*) or a hyphen (-) at the respective places.

\paragraph{Taverna/scufl} The Taverna Workflow Management System
\cite{taverna2013} is an open source scientific dataflow manager
developed by the mygrid~consortium in the UK.
%
Taverna includes a GUI-based rich-client workbench for workflow design and a
workflow enactment runtime. Taverna can enact workflows expressed in the scufl
language. Scufl is a near Turing complete dataflow language with well-defined
semantics \cite{turi-missier-etal:2007,glatard-montagnat:2008}.
